---
title: "Data Analysis for Prosper Loan Data Set"
author: "Lukas Stylianou"
date: "January 25, 2019"
output: 
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    theme: lumen
    highlight: textmate
---


```{r echo=FALSE, message=FALSE, warning=FALSE, packages}

library(ggplot2)
library(gridExtra)
library(GGally)
library(dplyr)
library(scales)
library(memisc)

# knitr: Suppress code/messages/warnings 
knitr::opts_chunk$set( echo=FALSE,warning=FALSE,message=FALSE)

# Set default plot options and center them
knitr::opts_chunk$set(fig.width=9,fig.height=5,fig.path='Figs/',
                      fig.align='center',tidy=TRUE,
                      echo=FALSE,warning=FALSE,message=FALSE)

# Don't use scientific notations for numbers
options(scipen = 999)

```

```{r echo=FALSE, Load_the_Data}
# Load the Data
df <- read.csv('prosperLoanData.csv')
# head(df, 50)
```

This report explores the Prosper Loan data set which contains 113,937 loans with
81 variables on each loan, including loan amount, borrower rate (or interest 
rate), current loan status, borrower income, borrower employment status, 
borrower credit history, and the latest payment information.

# Univariate Plots Section

```{r}
dim(df)
```


```{r}
str(df)
```

## Info About Loans
Let's look at the distribution of loans according to their terms in months.

```{r Loan terms, message=FALSE, warning=FALSE}

table(df$Term)

# Simple bar plot
ggplot(aes(x = as.factor(Term)), data = df) +
  geom_bar(color = 'black')

```

It is also interesting to see, how high the loan amounts are (with different bin 
sizes \$1000 and \$5000). Most loans are under \$5000.

```{r}

# Compare Loan amounts in hists with different bin sizes
amt1 <- ggplot(aes(x = LoanOriginalAmount), data = df) +
  geom_histogram(binwidth = 1000, breaks = seq(0, 30000, 1000), color = 'black')
amt2 <- ggplot(aes(x = LoanOriginalAmount), data = df) +
  geom_histogram(binwidth = 5000, breaks = seq(0, 30000, 5000), color = 'black')
grid.arrange(amt1, amt2, ncol = 1)

```


Loans range from \$1000 to \$35000 max. The median is \$6500.

```{r}

summary(df$LoanOriginalAmount)

```


Applying log scale on the histogram does not make any signs of normality visible
.

```{r}

# Log scale on loan amt
ggplot(aes(x = LoanOriginalAmount), data = df) +
  geom_histogram(bins = 11, color = 'black') +
  scale_x_log10()

```

Distribution of Borrower annual rates shows that most rates range from ~0.15% to
~0.2%.

```{r warning=FALSE}

# Overall comparison between Rates
p1 <- ggplot() + 
  geom_histogram(aes(x=BorrowerAPR, y=..count.., fill="grey", colour="black"), 
                 alpha=.6, data=df, stat = "bin", 
                 binwidth = .025, boundary = 0) +
  scale_colour_manual(name="Counts", values=c("grey" = "grey", "black"="black"),
                      labels=c("black"="BorrowerAPR", "grey"="BorrowerAPR")) +
  scale_fill_manual(name="Counts", values=c("grey" = "grey", "black"="black"), 
                    labels=c("black"="BorrowerAPR", "grey"="BorrowerAPR"))
  
p2 <- ggplot() + 
  geom_histogram(aes(x=BorrowerRate, y=..count.., fill="r", colour="r"), 
                 alpha=.3, data=df, stat = "bin", 
                 binwidth = .025, boundary = 0) +
  geom_histogram(aes(x=LenderYield,y=..count.., fill="b", colour="b"), alpha=.3,
                 data=df, stat = "bin", binwidth = .025, boundary = 0) +
  scale_colour_manual(name="Counts", values=c("r" = "red", "b"="blue"), 
                      labels=c("b"="LenderYield", "r"="BorrowerRate")) +
  scale_fill_manual(name="Counts", values=c("r" = "red", "b"="blue"), 
                    labels=c("b"="LenderYield", "r"="BorrowerRate")) +
  xlab('LenderYield | BorrowerRate')
  
grid.arrange(p1, p2, ncol = 1)

```

Loans per year/month
It turns out that date is in the date format. We will extract years and months
to create plots for time-series and seasonality. Most loans were taken in 2013.
Also it turns out, that the number of loans increases with the course of the
year and that most people take out a loan in October and December.
It would be interesting to analyse how high the total amount of all loans was
per year and what the average is. Year 2014 was removed from the monthly 
perspective, as it was not finished completely when the dataset was created and 
there would #todo: to univariate
Average loan amounts per year. And total amounts. It is well possible, that 2014
is not over yet, as the sum of loans is not quite as high as 2013, however, the
average is so far still higher than 2013.

```{r message=FALSE, warning=FALSE}

# Extract Year, Month
df$Year <- format(as.Date(df$LoanOriginationDate, 
                          format="%Y-%m-%d %H:%M:%S"),"%Y")
df$Year <- as.numeric(df$Year)
df$Month <- format(as.Date(df$LoanOriginationDate, 
                           format="%Y-%m-%d %H:%M:%S"),"%m")

# Compare Yearly and Monthly impacts on loans
t1 <- ggplot(aes(x = Year), data = df) +
  geom_bar() +
  scale_x_continuous(breaks = seq(2005, 2014))
  geom_histogram(stat = 'count')

t2 <- ggplot(aes(x = Month), data = subset(df, Year < 2014 & Year > 2006)) +
  geom_histogram(stat = 'count', binwidth = 1)

df.yearly <- df %>%
  group_by(Year) %>%
  summarise(mean_amt = mean(LoanOriginalAmount),
            median_amt = median(LoanOriginalAmount),
            min_amt = min(LoanOriginalAmount),
            max_amt = max(LoanOriginalAmount),
            sum_amt = sum(LoanOriginalAmount),
            n = n()) %>%
  arrange(Year)


t3 <- ggplot(data = df.yearly) +
  geom_bar(aes(x = Year, y = mean_amt), stat = 'identity') +
  scale_x_continuous(breaks = seq(2005,2014,1))

t4 <- ggplot(data = df.yearly) +
  geom_bar(aes(x = Year, y = sum_amt), stat = 'identity') +
  scale_x_continuous(breaks = seq(2005,2014,1))

grid.arrange(t1, t4, t3, t2, ncol = 1)

```

The bulk of the distribution of the prosper score lies in the middle. Most of
people get average scores.

```{r warning=FALSE}

# Prosper score hist
ggplot(aes(x = ProsperScore), data = df) +
  geom_histogram(bins = 11, color = 'black') +
  scale_x_continuous(breaks = seq(1,11,1))

```

Taking a look at the usage categories of loans, it turns out that more than 50%
are user for debt consolidation.

```{r}

lab <- c('Debt Consolidation', 
       'Home Improvement',
       'Business',
       'Personal Loan',
       'Student Use',
       'Auto',
       'Baby&Adoption',
       'Boat',
       'Cosmetic Procedure',
       'Engagement Ring',
       'Green Loans',
       'Household Expenses',
       'Large Purchases',
       'Medical/Dental',
       'Motorcycle',
       'RV',
       'Taxes',
       'Vacation',
       'Wedding Loans',
       'Other',
       'Not Available')

# Create new columns with text categories
df$ListingCategory <- factor(df$ListingCategory..numeric., 
                            levels = c(1:6,8:20,7,0), labels = lab)

ggplot(aes(x = df$ListingCategory, ..count.. / nrow(df)), data = df) +
  geom_bar() +
  scale_y_sqrt() +
  theme(axis.text.x = element_text(angle = 0, hjust = 1)) +
  xlab("Listing Category") +
  ylab("Count") +
  coord_flip()

```

## Info About Borrowers
Most of the recipients are employed, however they also do not specify on the
type of their emplyoment. It would be interesting to find out, how long
employment status last depending on each type. There are many empty employment
types which were changed to 'Not available' in order to clean the data.

```{r}

df$EmploymentStatus[df$EmploymentStatus == ''] <- 'Not available'
ggplot(aes(x = EmploymentStatus), data = df) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  geom_bar()

```

The incomes do not seem to be ordered. This needs to be fixed. Also, there is a
group labeled as 'Not employed'. One has to ask, whether to join this group with
the $0 group, however, it can not be said whether the 'Not employed' group can
be treated as such, as they may have other income sources (e.g. stocks, rental 
income, ...)

```{r}

table(df$IncomeRange)

```

Looking at the distribution of income ranges, most people earn between \$25000
and \$49999. The second larges group ranges from \$50000 to \$74999. It would
be interesting to see, how the different income ranges are composed in terms
of employment status, which we will see in the multi-variate part of the
analysis.

```{r IncomeRange, warning=FALSE}

df$IncomeRangeOrdered <- ordered(df$IncomeRange, levels = c("Not employed",
                                                            "$0",
                                                            "$1-24,999",
                                                            "$25,000-49,999",
                                                            "$50,000-74,999",
                                                            "$75,000-99,999",
                                                            "$100,000+",
                                                            "Not displayed"))
ggplot(aes(x = IncomeRangeOrdered), data = df) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  geom_histogram(stat = 'count')

```

```{r warning=FALSE}

p1 <- ggplot() + 
  geom_histogram(aes(x=BorrowerAPR, y=..count.., fill="grey", colour="black"), 
                 alpha=.6, data=df, stat = "bin", 
                 binwidth = .025, boundary = 0) +
  scale_colour_manual(name="Counts", values=c("grey" = "grey", "black"="black"),
                      labels=c("black"="BorrowerAPR", "grey"="BorrowerAPR")) +
  scale_fill_manual(name="Counts", values=c("grey" = "grey", "black"="black"),
                    labels=c("black"="BorrowerAPR", "grey"="BorrowerAPR"))


d1 <- ggplot() +
  geom_histogram(aes(x = CurrentDelinquencies, y=..count.., fill = 'red', 
                     colour = 'red'), alpha = .2, stat = "bin", 
                 data = df, boundary = 0, binwidth = 5) +
  geom_histogram(aes(x = DelinquenciesLast7Years, y=..count.., fill = 'blue', 
                     colour = 'blue'), alpha = .2, data = df, boundary = 0, 
                 binwidth = 5) +
  scale_y_log10(breaks = c(10, 100, 1000, 10000)) +
  scale_colour_manual(name="Counts", values=c("red" = "red", "blue"="blue"), 
    labels=c("blue"="DelinquenciesLast7Years", "red"="CurrentDelinquencies")) +
  scale_fill_manual(name="Counts", values=c("red" = "red", "blue"="blue"), 
    labels=c("blue"="DelinquenciesLast7Years", "red"="CurrentDelinquencies"))

d2 <- ggplot(data = df) +
  geom_histogram(aes(x = PublicRecordsLast10Years, color = 'blue', 
                     fill = 'blue'), alpha = .2, boundary = 0, binwidth = 1) +
  geom_histogram(aes(x = PublicRecordsLast12Months, color = 'red', 
                     fill = 'red'), alpha = .2, boundary = 0, binwidth = 1) +
  scale_x_continuous(limits = c(0, 20)) +
  scale_y_log10(breaks = c(10, 100, 1000, 10000)) +
  scale_colour_manual(name="Counts", values=c("red" = "red", "blue"="blue"), 
                      labels=c("blue"="PublicRecordsLast10Years", 
                               "red"="PublicRecordsLast12Months")) +
  scale_fill_manual(name="Counts", values=c("red" = "red", "blue"="blue"), 
                    labels=c("blue"="PublicRecordsLast10Years", 
                             "red"="PublicRecordsLast12Months"))

grid.arrange(d1, d2, ncol = 1)

```

Open vs Current Credit lines
Open and current credit lines are a right-skewed distribution.

```{r warning=FALSE}

ggplot( data = df) +
  geom_histogram(aes(x = CurrentCreditLines, fill = 'blue'), 
                 alpha = .3, binwidth = 2, boundary = 0) +
  geom_histogram(aes(x = OpenCreditLines, fill = 'red'), 
                 alpha = .3, binwidth = 2, boundary = 0) +
  coord_cartesian(xlim = c(0,40)) +
  scale_x_continuous(breaks = seq(0, 40, 2)) +
  scale_colour_manual(name="Counts", values=c("red" = "red", "blue"="blue"),
                      labels=c("blue"="CurrentCreditLines",
                               "red"="OpenCreditLines")) +
  scale_fill_manual(name="Counts", values=c("red" = "red", "blue"="blue"), 
                    labels=c("blue"="CurrentCreditLines", 
                             "red"="OpenCreditLines"))

```


Home ownership is around 50-50 (Slightly more home owners)

```{r}

summary(df$IsBorrowerHomeowner)

```

Most borrowers have high bank card utilization. Utilisation over 1.0 would then
mean overdrafting the bank account.

```{r message=FALSE, warning=FALSE}

# Bankcard utilisation bins
ggplot(aes(x = BankcardUtilization), data = df) +
  geom_histogram(binwidth = .2, color = 'black', boundary = 0) +
  scale_x_continuous(limits = c(0, 1.5))

```

Most people take loans as individuals.

```{r message=FALSE, warning=FALSE}

# Simple True False hist
ggplot(aes(x = CurrentlyInGroup), data = df) +
  geom_histogram(stat = 'count')

```

# Univariate Analysis

### What is the structure of your dataset?

The dataset contains 113937 observations of 81 features.

### What is/are the main feature(s) of interest in your dataset?

BorrowerAPR

BorrowerRate

LenderYield

ProsperScore

EmploymentStatus

EmploymentStatusDuration

BankcardUtilization

IncomeRange

LoanOriginalAmount

LoanOriginationDate

Term


### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?

ListingCategory

IsBorrowerHomeowner

CurrentlyInGroup

CurrentDelinquencies

DelinquenciesLast7Years

PublicRecordsLast12Months

PublicRecordsLast10Years

AvailableBankcardCredit

CurrentCreditLines

OpenCreditLines


### Did you create any new variables from existing variables in the dataset?
Year

Month


### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?
Employment Status had an empty values issue which is why they were changed to 
"Not available" to match the existing variable.
As Income Range was unordered, the ordering was adjusted and 100k+ incomes were
put to the right position.
Also, the date column was unusable which is why Year and Month were extracted
and put into separate columns.
Listing Categories were converted from numerical values to strings, in order
to understand the different categories better in the plot.

# Bivariate Plots Section

As "Not employed" is much steeper than Employed and e.g. full time this suggests
that most people are fortunately unemployed only for a relatively short time.
Also part time employment does not seem to last as long as full time, having 
bulk of distribution closer to the right.

```{r echo=FALSE, message=FALSE, warning=FALSE}

# wrap Duration & Density over Employment Status 
ggplot(aes(x = EmploymentStatusDuration, y = ..density.. ), 
       data = subset(df, EmploymentStatus != 'Not available')) +
  geom_freqpoly(binwidth = 30) +
  scale_x_continuous(limits = c(0, 365)) +
  facet_wrap(~ EmploymentStatus)
```

The borrower annual rate decreases with the increase in their score. This is
also a moderate correlation of 0.65.

```{r message=FALSE, warning=FALSE}

# Analyse relationship between Score and APR
cor.test(df$ProsperScore, df$BorrowerAPR)

ggplot(aes(x = ProsperScore, y = BorrowerAPR), data = df) +
  geom_jitter(alpha = .02) +
  coord_cartesian(ylim = c(0.05, 0.4)) +
  scale_x_continuous(breaks = seq(1,11,1)) +
  geom_smooth()

```

Borrower Rate (obviously) has highly linear relation to Lender yield, however, there are
some values, which are not on the line... We will analyse this phenomenon in the
multi variate analysis.

```{r}

# Analyse relation between Rate and Yield
cor.test(df$BorrowerRate, df$LenderYield)

ggplot(aes(x = BorrowerRate, y = LenderYield), data = df) +
  geom_jitter(alpha = .02) +
  coord_cartesian(xlim = c(0.05, 0.4), ylim = c(0.05, 0.4))

```

Thers is slight correlation between loan amount and and APR.

```{r}

cor.test(df$BorrowerAPR, df$LoanOriginalAmount)

```

```{r message=FALSE, warning=FALSE}

# Relation between APR and Amount
ggplot(aes(y = BorrowerAPR, x = LoanOriginalAmount), data = df) +
  geom_jitter(alpha = .03, size = 1) +
  geom_smooth()

```

Mean APR had its highest point in 2011 and then decreased again.

```{r message=FALSE, warning=FALSE}

# Yearly trend of rates
cor.test(df$Year, df$BorrowerAPR)

df.apr <- df %>%
  group_by(Year) %>%
  summarise(mean_apr = mean(BorrowerAPR),
            median_apr = median(BorrowerAPR),
            min_apr = min(BorrowerAPR),
            max_apr = max(BorrowerAPR),
            n = n()) %>%
  arrange(Year)

ggplot(data = df.apr) +
  geom_line(aes(x = Year, y = mean_apr)) +
  geom_line(aes(x = Year, y = min_apr), linetype = 2, color = 'grey') +
  geom_line(aes(x = Year, y = max_apr), linetype = 2, color = 'grey') +
  scale_x_continuous(limits = c(2007, 2014), breaks = seq(2007, 2014, 1))

```

Most borrowers have less than 50 Credit lines and less than 25 delinquencies in 
the past 7 years.

```{r message=FALSE, warning=FALSE}

# Ratio of CreditLines and Delinquencies
ggplot(aes(x = TotalCreditLinespast7years, y = DelinquenciesLast7Years), data = df) +
  geom_point(alpha = .05)

```

People with higher bakcard utilisation tend to have higher loan rates.

```{r message=FALSE, warning=FALSE}

# Relation between Banckard utilisation and APR
cor.test(df$BankcardUtilization, df$BorrowerAPR)

ggplot(aes(x = BankcardUtilization, y = BorrowerAPR), data = df) +
  geom_point(alpha = .01) +
  scale_x_continuous(limits = c(0,1.5)) +
  geom_smooth()

```

People with more income tend to borrow higher amounts. Only people earning over
100k got loans higher than 25k.


```{r message=FALSE, warning=FALSE}

# Plot Income Ranges and Loan Amount incl. mean
ggplot(aes(x = IncomeRangeOrdered, y = LoanOriginalAmount), data = df) +
  geom_boxplot() +
  stat_summary(fun.y = mean, geom = 'point', shape = 4) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

Borrowers with higher income also get better anual rates. The fact that people
with an income of $0 get low rates could be that there are many students in this
group who get cheap student loans.

```{r message=FALSE, warning=FALSE}

# Plot Income ranges and APR including mean
ggplot(aes(x = IncomeRangeOrdered, y = BorrowerAPR), data = df) +
  geom_boxplot() +
  stat_summary(fun.y = mean, geom = 'point', shape = 4) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

Prosper Score depends on income. If score is smaller than 50k, it stays the same
on average. It is also the same for ranges from 50k to 75k. Average is
highest above 100k.

```{r message=FALSE, warning=FALSE}

# Plot income range and score including mean
ggplot(aes(x = IncomeRangeOrdered, y = ProsperScore), data = df) +
  geom_boxplot() +
  stat_summary(fun.y = mean, geom = 'point', shape = 4) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

On log scale, delinquencies become better visible. However, they do not really
vary visibly among income ranges.

```{r message=FALSE, warning=FALSE}

# Plot income range and delinquencies incl. mean
ggplot(aes(x = IncomeRangeOrdered, y = DelinquenciesLast7Years), data = df) +
  scale_y_log10() +
  geom_boxplot() +
  stat_summary(fun.y = mean, geom = 'point', shape = 4) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```



# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

It turns out that people with higher income on average get higher amounts and
better rates for loans.

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

Loan rates increased until 2011 and then decreased again. Maybe this is a
consequence of the financial crisis around 2008.

### What was the strongest relationship you found?
BorrowerRate, LenderYield -> 0.99
BorrowerAPR, ProsperScore -> -0.67
ProsperScore, LoanOriginalAmount -> -0.32

# Multivariate Plots Section

As "Not employed" and "$0" are not of the same quality, the analysis of 
employment statuses was conducted to see the composition of different statuses.

```{r message=FALSE, warning=FALSE}

# Look at composition of different Income Ranges in terms of employment status
ggplot(aes(x = IncomeRangeOrdered, fill = EmploymentStatus), data = df) +
  geom_histogram(stat = 'count') +
  scale_y_log10(labels = scales::scientific_format()) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

Borrower rate is strictly linear to the Lender Yield. However, if there
is no score available, the lender yield seems to deviate from the ideal margin 
line as can be seen in the following two plots. The upper one has the NA Prosper
Score values removed while the lower one incorporates them as grey dots.

```{r}

# relation between rate and yield with AND without NA data
n1 <- ggplot(aes(x = BorrowerRate, y = LenderYield, color = ProsperScore), data = subset(df, !is.na(ProsperScore))) +
  geom_jitter(alpha = .02) +
  coord_cartesian(xlim = c(0.05, 0.4), ylim = c(0.05, 0.4))

n2 <- ggplot(aes(x = BorrowerRate, y = LenderYield, color = ProsperScore), data = df) +
  geom_jitter(alpha = .02) +
  coord_cartesian(xlim = c(0.05, 0.4), ylim = c(0.05, 0.4))

grid.arrange(n1, n2, ncol = 1)

```

Relationship between BorrowerAPR, Loan Amount and Score shows that APR slightly 
decreases when amount increases and when score increases. 

```{r}

# Scatterplot - Original Loan Amount and Borrow APR, by Prosper Score
ggplot(aes(x = LoanOriginalAmount, y = BorrowerAPR, color = ProsperScore),
       data = subset(df, !is.na(ProsperScore))) +
  geom_jitter(alpha = .1) +
  scale_color_gradient2("Prosper\nScore",
        midpoint = median(subset(df,!is.na(ProsperScore))$ProsperScore))

```

Long term loans (60 months) tend to have higher amounts than shorter loans (36 
months). Short term loans (12 tend to have the lowest amounts).

```{r echo=FALSE, Multivariate_Plots}

# look at term based on amount & apr
ggplot(aes(x = LoanOriginalAmount, y = BorrowerAPR, color = as.factor(Term)),
       data = subset(df, !is.na(ProsperScore))) +
  geom_jitter(alpha = 1) +
  scale_color_brewer(palette = 'Set1')

```

When comparing Income Ranges in terms of loan amounts, it can be seen that the
bulks of the high income ranges are more to the right side (higher loans) and 
the low income ranges on the left side (lower loans).

```{r}

# Density - Original Loan Amount, by Income Range (Log 10  scale)
ggplot(aes(x = LoanOriginalAmount, y = ..density.., color = IncomeRangeOrdered),
       data = subset(df, IncomeRangeOrdered != "Not employed" & 
                       IncomeRangeOrdered != "Not displayed")) +
  scale_x_log10() +
  geom_freqpoly(bins = 8) +
  scale_color_brewer(palette = "RdYlBu", direction = -1, name = "Income\nRange") +
  theme_dark()

```


Bank card utilisation per income range shows that higher income ranges have a 
larger bulk at high credit card utilisation rates.

```{r message=FALSE, warning=FALSE}

# density distribution of bankcard utilisation wrapped by income
ggplot(aes(x = BankcardUtilization, y = ..density..), 
       data = subset(df, !is.na(IncomeRangeOrdered))) +
  geom_freqpoly(binwidth = .1) +
  scale_x_continuous(limits = c(0,1.5)) +
  facet_wrap(~ IncomeRangeOrdered)

```

Mean loan amount per year per income range again clearly shows, that higher
income ranges take get higher loans. Also it can be seen, that there is no
value for $0 for 2014. The group is not very realistic, which may be why it was
dropped. "Not displayed" shows a lack of data before 2007.

```{r}

# yearly trend of mean amounts by income
ggplot(aes(x = Year, y = LoanOriginalAmount), data = df) +
  geom_line(stat = 'summary', fun.y = mean, aes(color = IncomeRangeOrdered)) +
  scale_x_continuous(breaks = seq(2005,2014,1))

```

Relationship between borrower rate, term and its APR Different Terms
are clustered in this plot based on their Rate/APR ratio.

```{r message=FALSE, warning=FALSE}

# borrower rate by apr colored by term
ggplot(aes(x = BorrowerRate, y = BorrowerAPR, color = Term), data = df) +
  geom_point()

```

Lower income is connected to lower employment duration.
```{r message=FALSE, warning=FALSE}

# income by employment status density
df.empl <- subset(df, EmploymentStatus == 'Employed' | 
                    EmploymentStatus == 'Full-time' | 
                    EmploymentStatus == 'Part-time' | 
                    EmploymentStatus == 'Self-employed' )
df.empl <- subset(df.empl, IncomeRangeOrdered != "Not employed" & 
                    IncomeRangeOrdered != "Not displayed")

ggplot(aes(x = EmploymentStatusDuration, y = ..density.., color = IncomeRange), 
       data = df.empl) +
  scale_x_log10() +
  geom_freqpoly(bins = 12)

```


Lower income ranges have much higher APRs. After 2012 they even increased,
while for higher income ranges APRs decreased and ceased to exist after 2013.

```{r}

# yearly change of apr by income (color)
ggplot(aes(x = Year, y = BorrowerAPR, color = IncomeRangeOrdered), 
       data = subset(df, IncomeRangeOrdered != 'Not displayed')) +
  geom_line(stat = 'summary', fun.y = mean) +
  scale_x_continuous(limits = c(2007, 2014), breaks = seq(2007, 2014, 1))

```

Linear Model for score

```{r}

# build linear model (multinomial)
m1 <- lm(BorrowerAPR ~ ProsperScore, data = df)
m2 <- update(m1, ~ . + LoanOriginalAmount)
m3 <- update(m2, ~ . + Year)
m4 <- update(m3, ~ . + EmploymentStatus)
m5 <- update(m4, ~ . + BankcardUtilization)
m6 <- update(m5, ~ . + IncomeRange)
m7 <- update(m6, ~ . + Month)
m8 <- update(m7, ~ . + Term)
m9 <- update(m8, ~ . + CurrentDelinquencies)
m10 <- update(m9, ~ . + DebtToIncomeRatio)
m11 <- update(m10, ~ . + CurrentlyInGroup)
m12 <- update(m11, ~ . + IsBorrowerHomeowner)
mtable(m1, m2, m3, m4, m5, m6, m7, m8, m10, m11, m12)

```


# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

Conducting multivariate analyses it becomes even clearer that based on how high
the income is, the loans are higher and have lower rates. The plots also
visualise how the rates and amounts changed over the years.
It could also be credit card utilisation patterns vary within different income 
groups.

### Were there any interesting or surprising interactions between features?
Prosper did not differentiate before 2007 and does not have any borrowers with
zero income on unemployment status after as of 2013 (maybe because they did not
allow any loans for this group.) 


### OPTIONAL: Did you create any models with your dataset? Discuss the \
strengths and limitations of your model.

The multinomial linear regression model takes into account 12 different 
variables and achieves an R^2 of 0.645 for predicting BorrowerAPR. It only looks
at linear relationships of variables. Therefore, performance of the model could 
be improved, by looking at non-linear relations between variables. Also,
advanced feature engineering and cleaning of the data set and adding more 
variables could improve performance.

------

# Final Plots and Summary

### Plot One

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Number of Loans by Month (removed incomplete years)
ggplot(aes(x = Month), data = subset(df, Year < 2014 & Year > 2006)) +
  geom_histogram(stat = 'count', binwidth = 1) +
  xlab("Month") +
  ylab ("Number of Loans taken out (n)") +
  ggtitle("Monthly Distribution of Loans")

```

### Description One
Most loans are being taken out during the end of the year. This is quite
interesting and can have several reasons. The most intuitive one is that people
have plans for the following year and therefore borrow money. Another theory
might be that people run out of money during the end of the year and need to
borrow.

### Plot Two
```{r echo=FALSE, message=FALSE, warning=FALSE}

# Distribution of BorrowerAPR, by Income Range
# bps are basic points
ggplot(aes(x = IncomeRangeOrdered, y = BorrowerAPR), data = df) +
  geom_boxplot() +
  stat_summary(fun.y = mean, geom = 'point', shape = 4, size = 3,
               aes(color = "Mean")) +
  scale_color_manual('', values = c("Mean" = 'red')) +
  theme(axis.text.x = element_text(angle = 0, hjust = 1)) +
  xlab("Income Range ($)") +
  ylab("Borrower APR (bps)") +
  ggtitle("Distribution of BorrowerAPR, by Income Range") +
  coord_flip()

```

### Description Two
The BorrowerAPR is highly dependent on the income range. Having more income
makes it more likely to get a better rate. On the other hand, having lower
income increases the chances to get a worse rate. There is an exception with
earning $0. This group could include students who get cheaper student loans.

### Plot Three
```{r echo=FALSE, message=FALSE, warning=FALSE}

# 
ggplot(aes(x = Year, y = BorrowerAPR * 100, color = IncomeRangeOrdered), 
       data = subset(df, IncomeRangeOrdered != 'Not displayed')) +
  geom_line(stat = 'summary', fun.y = mean) +
  scale_x_continuous(limits = c(2007, 2014), breaks = seq(2007, 2014, 1)) +
  xlab("Year") +
  ylab("BorrowerAPR [%]")

```

### Description Three
This plot shows the chronological sequence of the APR based on the borrowers'
income ranges. It can be seen that the overall rate increases until 2011 and
subsequently drops, except for low earners and unemployed borrowers who cease
to exist in the data set as of 2013.


------

# Reflection

This data analysis included the behaviour of various borrowers based on
various features. Insight was gained, especially considering the borrowers'
employment statuses, income ranges, credit card utilisation and use of loans.
Finally, a model was generated to predict borrowers rates based on 12 of their
characteristics.

The dataset was challenging to some extent. As it includes approx. 80 variables
a lot of work was initially pu into understanding the dataset and looking at 
different variables. Ggpairs on a short list of features was especially helpful
and also using the spreadsheet supported with getting a better grasp on the
different features.

Additionally, some of the struggles included having to deal with features which
were not internally congruent, e.g. income range, which included "Not employed",
even though this should have been part of "Employment Status" only. Also,
different values of "Employment Status" did not have the same quality. While,
"Employed" and "Unemployed" are contrary, "Full-time" and "Part-time" are 
subsets of "Employed" and cannot be directly compared with "Employed".

On the other hand, it was suprising to see, how the different features 
influence the BorrowerAPR (e.g. Score and Income Range). It was also suprising
that low-income and unemployed borrowers ceased to exist in the data set as of
2013.

Further cleaning, feature engineering and exploring more variables could be 
taken into account in order to gain further insights and build more powerful
models. For example, it could be explored, which estimated values (e.g. 
EstimatedLoss, EstimatedReturn, EstimatedYield ...) have on the model or what 
impact variables such as Occupation or CurrentlyInAGroup have. Also, it
could be analysed, if the performance of the model changes, when getting rid
of the pre-2007 values, which do not have different income ranges incorprated.
Further features could be engineered and used for prediction, e.g. ratios
between features such as Credit Lines by Delinquencies. Finally, it could be
analysed if non-linear assumptions or higher-range functions of features
generate better model results.